Looking at Non-Linear Dimension Reductions as Models in the Data Space

### 2025-06-17 paper.qmd
Removed text with "links" in captions

```{r}
#| label: best-fit-tsne
#| fig-cap: "Lifting the fitted model into \\pD{}. Starting from a default tSNE layout (a) of nonlinear clusters data ($n =  2000$ and $p = 4$), from which the \\gD{} wireframe is constructed (a) and then lifted to \\fD{} (b, c). The fit is reasonbly tight with the data, although it does not fully spread the full width. This highlights a key characteristic of tSNE, which compresses the data when applying tSNE. Furthermore, the sparse space observed in the middle of the tSNE layout is a result of (hyper-)~parameter choice. The link to the video of the langevitour animation is available in @tbl-links."
#| fig-width: 8
#| fig-height: 2

@@

```{r}
#| label: fig-five-gau-projs
#| fig-height: 10
#| fig-width: 15
#| fig-pos: H
#| fig-cap: "NLDR's organise points in the \\gD{} layout in different ways, possibly misleadingly, illustrated using three layouts: (a) tSNE, (b) UMAP, (c) PaCMAP. The data has five Gaussian cluster in \\fD{}. The bottom row of plots shows a \\gD{} projection from a tour on \\fD{} revealing the differences generated by the layouts on the model fits.  We would expect the model fit to be like that in (a2) where it is distinctly is separate for each cluster but like a hairball in each. This would indicate the distinct clusters, each being fully \\fD{}. With (c2), the curiousity is that the model is a \\gD{} pancake shape in \\fD{}, indicating that there is some ordering of points done by PaCMAP, posisbly along some principal component axes. The links to the videos of the langevitour animations are available in @tbl-links."

@@

```{r}
#| label: fig-one-dens_clust-error
#| fig-width: 15
#| fig-height: 10
#| fig-pos: H
#| out-width: 100%
#| fig-cap: "The sparse end of the structure shows a high model error, while the dense end shows a low model error. The tSNE layout of the C-shaped structure, with dense points (a1-a3) are colored according to their \\fD{} model error. Darker colors represent a high error, while lighter colors indicate a low error. a2, a3 and b2, b3 present the results from linked plots that brushed the high \\fD{} error points in \\gD{} and \\fD{}. It helps in identifying the high \\fD{} model errors that occur due to the sparse end. The link to the video of the linked plots is available in @tbl-links."

@@

```{r}
#| echo: false
#| fig-cap: "Compare the published \\gD{} layout (a) made with UMAP and the \\gD{} layout selected by RMSE plot (e) made by tSNE, both with default (hyper-)parameters. The two plots on the right show  projections from a tour, with the models overlaid. The published layout suggested three very separated clusters, but this is not present in the data. While there may be three clusters they are not well-separated. The difference in model fit also indicates this: the published layout a does not spread out fully into the point cloud like the model generated from layout e. This supports the choice that layout e is the better representation of the data, because it does not exaggerate separation between clusters."
#| label: fig-model-pbmc-author-proj
#| fig-pos: H
#| fig-width: 10
#| fig-height: 15
#| out-width: 90%

@@

```{r}
#| label: fig-clust-mnist
#| fig-pos: H
#| fig-height: 5
#| fig-width: 15
#| out-width: 100%
#| fig-cap: "The tSNE layout of the MNIST digit 1 data shows a big nonlinear cluster (grey) and a small cluster (orange) located very close to the one corner of the big cluster in \\gD{} (a). The MNIST digit 1 data ($n =  7877$ and $p = 10$) has a nonlinear structure in $10\\text{-}D$. Two \\gD{} projections from a tour on $10\\text{-}D$ reveal that the closeness of the clusters in $10\\text{-}D$ and the twisted pattern of the model fit with tSNE ($a_1 = 0.041$, $b = 659/585 \\text{ } (30, \\text{ }48)$). The link to the video of the langevitour animation is available in @tbl-links. The brushing feature in the linked plots helps in visualizing the closeness of the small cluster to the big cluster. The link to the video of the linked plots is available in @tbl-links."

>>>>

```{r}
#| label: best-fit-tsne
#| fig-cap: "Lifting the fitted model into \\pD{}. Starting from a default tSNE layout (a) of nonlinear clusters data ($n =  2000$ and $p = 4$), from which the \\gD{} wireframe is constructed (a) and then lifted to \\fD{} (b, c). The fit is reasonbly tight with the data, although it does not fully spread the full width. This highlights a key characteristic of tSNE, which compresses the data when applying tSNE. Furthermore, the sparse space observed in the middle of the tSNE layout is a result of (hyper-)~parameter choice."
#| fig-width: 8
#| fig-height: 2

@@

```{r}
#| label: fig-five-gau-projs
#| fig-height: 10
#| fig-width: 15
#| fig-pos: H
#| fig-cap: "NLDR's organise points in the \\gD{} layout in different ways, possibly misleadingly, illustrated using three layouts: (a) tSNE, (b) UMAP, (c) PaCMAP. The data has five Gaussian cluster in \\fD{}. The bottom row of plots shows a \\gD{} projection from a tour on \\fD{} revealing the differences generated by the layouts on the model fits.  We would expect the model fit to be like that in (a2) where it is distinctly is separate for each cluster but like a hairball in each. This would indicate the distinct clusters, each being fully \\fD{}. With (c2), the curiousity is that the model is a \\gD{} pancake shape in \\fD{}, indicating that there is some ordering of points done by PaCMAP, posisbly along some principal component axes."

@@

```{r}
#| label: fig-one-dens_clust-error
#| fig-width: 15
#| fig-height: 10
#| fig-pos: H
#| out-width: 100%
#| fig-cap: "The sparse end of the structure shows a high model error, while the dense end shows a low model error. The tSNE layout of the C-shaped structure, with dense points (a1-a3) are colored according to their \\fD{} model error. Darker colors represent a high error, while lighter colors indicate a low error. a2, a3 and b2, b3 present the results from linked plots that brushed the high \\fD{} error points in \\gD{} and \\fD{}. It helps in identifying the high \\fD{} model errors that occur due to the sparse end."

@@

```{r}
#| echo: false
#| fig-cap: "Compare the published \\gD{} layout (a) made with UMAP and the \\gD{} layout selected by RMSE plot (e) made by tSNE. The two plots on the right show  projections from a tour, with the models overlaid. The published layout suggested three very separated clusters, but this is not present in the data. While there may be three clusters they are not well-separated. The difference in model fit also indicates this: the published layout a does not spread out fully into the point cloud like the model generated from layout e. This supports the choice that layout e is the better representation of the data, because it does not exaggerate separation between clusters."
#| label: fig-model-pbmc-author-proj
#| fig-pos: H
#| fig-width: 10
#| fig-height: 15
#| out-width: 90%

@@

```{r}
#| label: fig-clust-mnist
#| fig-pos: H
#| fig-height: 5
#| fig-width: 15
#| out-width: 100%
#| fig-cap: "The tSNE layout of the MNIST digit 1 data shows a big nonlinear cluster (grey) and a small cluster (orange) located very close to the one corner of the big cluster in \\gD{} (a). The MNIST digit 1 data has a nonlinear structure in $10\\text{-}D$. Two \\gD{} projections from a tour on $10\\text{-}D$ reveal that the closeness of the clusters in $10\\text{-}D$ and the twisted pattern of the model fit with tSNE. The brushing feature in the linked plots helps in visualizing the closeness of the small cluster to the big cluster."
